---
title: "Gapminder"
author: "Pepper Wang"
date: 2026-10-6
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
?gapminder
summary(gapminder)
```

**Observations**:
- Variables:
  - Country
  - Continent
  - Year
  - Life expectancy, in years
  - Population
  - GDP per capita, in USD, inflation-adjusted

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`

years <- gapminder |> pull(year)

year_max <- max(years)
year_min <- min(years)

year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task-1}
## TASK: Create a visual of gdpPercap vs continent
gapminder |>
  filter(year == year_min) |>
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  scale_y_log10()
```

```{r q2-task-2}
## TASK: Create a visual of gdpPercap vs continent
gapminder |>
  filter(year == year_min) |>
  ggplot(aes(x = gdpPercap, fill = continent)) +
  geom_histogram(bins = 10, position = "dodge") +
  scale_x_log10()
```

```{r q2-task-3}
## TASK: Create a visual of gdpPercap vs continent
gapminder |>
  filter(year == year_min) |>
  ggplot(aes(x = gdpPercap, color = continent)) +
  geom_freqpoly(bins = 10) +
  scale_x_log10()
```

**Observations**:

- Africa has a large number of countries with low GDP per capita. Asia is similar but less so. Europe has a wide range that is in the middle. Oceania has very few countries but they all have very similar GDP per capita. The Americas have mostly countries with low GDP per capita, with some outliers that are higher.
- Asia has 1 country that has a GDP per capita many times higher than any other country.

**Difficulties & Approaches**:

- There is one extreme outlier country in Asia with a GDP per capita that is much higher than any other country. In every plot I tried using linear scaling, the outlier caused all the rest of the data to get squished and hard to read. Using log scaling makes this much better.

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder |>
  filter(year == year_min) |>
  arrange(desc(gdpPercap))
```

**Observations**:

- Identify the outlier countries from q2
  - The largest outlier is Kuwait. The two smaller outliers are Switzerland and the United States.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder |>
  filter(year == year_min | year == year_max) |>
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  scale_y_log10() +
  facet_grid(cols = vars(year)) +
  geom_point(
    data = . %>% filter(country %in% c(
      "Kuwait", "Switzerland", "United States"
    )),
    mapping = aes(color = country),
    size = 2
  )
```

**Observations**:

- Write your observations here
  - Every continent had generally much higher and more varied GDP per capita in 2007 compared to 1952.
  - Kuwait was a large outlier in 1952; in 2007, it is no longer such an extreme outlier, but it still has the highest GDP per capita in Asia.

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
gapminder |>
  filter(year == year_min | year == year_max) |>
  ggplot(aes(x = gdpPercap, y = lifeExp, color = as.factor(year))) +
  geom_point() +
  scale_x_log10()
```

- There seems to be a positive correlation between GDP per capita and life expectancy, though there is a lot of variation within it
- Between 1952 and 2007, overall GDP per capita and life expectancy both increased substantially, though the variation in both of these seems to have stayed about the same

```{r q5-task2}

gapminder |>
  filter(year == year_min | year == year_max) |>
  ggplot(aes(lifeExp, color = continent)) +
  geom_freqpoly(bins = 18) +
  facet_grid(rows = vars(year))
```

- Between 1952 and 2007, most continents' life expectancy had an overall increase
- Specifically, Asia and the Americas were previously in the low to medium range, and they both ended up on the high end of the plot, close to Europe. 

```{r q5-task3}
gapminder |>
  filter(year == year_min | year == year_max) |>
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot(aes(fill = continent)) +
  facet_grid(cols = vars(year))
```

- This graph conveys similar information to the previous one, about changes in life expectancy across continents. 

```{r q5-task4}
gapminder |>
  filter(country %in% c(
    "Kuwait", "Switzerland", "United States"
  )) |>
  ggplot(aes(x = year, y = gdpPercap, color = country)) +
  geom_line()
```

- Kuwait (the biggest GDP per capita outlier in 1952) has a GDP per capita that decreases from 1957 to 1987, until it is similar to the GDP per capita of Switzerland and the United States, then increases at about the same rate as Switzerland and the United States. 

